---
title: "Obj2.2: Nonparametric Model"
author: "Leonardo Leal Filho"
date: '2022-05-29'
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r}
library(tidyverse)
library(WDI)
library(countrycode)
library(GGally)
library(class)
```

```{r}
#Change working directory to this source file directory
setwd(dirname(rstudioapi::getSourceEditorContext()$path))
```

```{r}
LifeExpecFilePath = "../Datasets/LifeExpectancyData.csv"
df <- read.csv(LifeExpecFilePath)
```

## Checking the data

```{r}
dim(df)
summary(df)
```
## Checking the features with NA values and how many NA values each has

```{r}
# Creating an empty vector to collect the features that have NA values
na.feat <- c()

# Looping through the dataset to collect the features with NA values
for (i in names(df)) {
  if (sum(is.na(df[,i])) > 0) {
    na.feat <- c(na.feat, i)
  }
}

# Making an empty dataframe to store the names of the features with NA values and how many NA values they have
na.count <- data.frame(Features = na.feat, Count = numeric(length(na.feat)))

# Looping through the dataset to collect the count of NA values those features have 
for (i in 1:length(na.feat)) {
  na.count[i, 2] <- sum(is.na(df[, na.feat[i]]))
}

# Checking the na.count dataframe
na.count
```

## Removing some rows with NA values

```{r}
# Finding the rows of "thinnes..1.19.years" that contains NA values
n1 <- which(is.na(df$thinness..1.19.years))
# Removing the rows
df <- df[-n1,]
```

```{r}
# Creating an empty vector to collect the features that have NA values
na.feat <- c()

# Looping through the dataset to collect the features with NA values
for (i in names(df)) {
  if (sum(is.na(df[,i])) > 0) {
    na.feat <- c(na.feat, i)
  }
}

# Making an empty dataframe to store the names of the features with NA values and how many NA values they have
na.count <- data.frame(Features = na.feat, Count = numeric(length(na.feat)))

# Looping through the dataset to collect the count of NA values those features have 
for (i in 1:length(na.feat)) {
  na.count[i, 2] <- sum(is.na(df[, na.feat[i]]))
}

# Checking the na.count dataframe
na.count
```


```{r}
# Finding the rows of "Polio" that contains NA values
n1 <- which(is.na(df$Polio))
# Removing the rows
df <- df[-n1,]
```

```{r}
# Creating an empty vector to collect the features that have NA values
na.feat <- c()

# Looping through the dataset to collect the features with NA values
for (i in names(df)) {
  if (sum(is.na(df[,i])) > 0) {
    na.feat <- c(na.feat, i)
  }
}

# Making an empty dataframe to store the names of the features with NA values and how many NA values they have
na.count <- data.frame(Features = na.feat, Count = numeric(length(na.feat)))

# Looping through the dataset to collect the count of NA values those features have 
for (i in 1:length(na.feat)) {
  na.count[i, 2] <- sum(is.na(df[, na.feat[i]]))
}

# Checking the na.count dataframe
na.count
```
lastly I will remove the rows that contains missing values for the "Adult.Mortality" feature.

```{r}
# Finding the rows of "Adult.Mortality" that contains NA values
n1 <- which(is.na(df$Adult.Mortality))
# Removing the rows
df <- df[-n1,]
```

```{r}
# Creating an empty vector to collect the features that have NA values
na.feat <- c()

# Looping through the dataset to collect the features with NA values
for (i in names(df)) {
  if (sum(is.na(df[,i])) > 0) {
    na.feat <- c(na.feat, i)
  }
}

# Making an empty dataframe to store the names of the features with NA values and how many NA values they have
na.count <- data.frame(Features = na.feat, Count = numeric(length(na.feat)))

# Looping through the dataset to collect the count of NA values those features have 
for (i in 1:length(na.feat)) {
  na.count[i, 2] <- sum(is.na(df[, na.feat[i]]))
}

# Checking the na.count dataframe
na.count
```


```{r}
dim(df)
```
There are now 2888 observations left in the dataset, a loss of about 1.7% of the original amount of observations.  The goal from now on is to find ways to input the missing values to the dataset.

The population feature, which is the feature that gives us the information of the country's total population count, is the feature that has the most missing values (NAs).  Since the information comes per country per year, one way to solve this issue is to use the World Bank Database package, which contains this information as well as the GDP information, Hepatitis B and Total Expenditure.

I will start by acquiring the population data, but the actual first step will be to input the abreviation of the countries in a new column to make it easier to grab the other missing information.


```{r}
na.count
```
### I will now remove the features that have missing data left.  

### I will also remove the "Country" feature because since 50 rows were removed from the dataset, some countries may not be fully represented, if at all.  This means making prediction based on countries may not yield any results due to the country not being in the training of the model.

### The "Year" feature will also be removed because I am not planning to build a time series model.  Using the "Year" feature would mean that any prediction would be for a Year that is not within the actual dataset, which means the model would not have been trained on that value so it would extrapolate.
```{r}
# Removing the features that have 160 or more NA values plus the "Country" and "Year" features.
df2 <- df %>% select(!c(na.count$Features, "Country", "Year"))
```


```{r}
dim(df2)
str(df2)
```
### The only non numerical feature left is the "Status" feature.  Let me check how many factors are in that feature and what are they.


```{r}
# Transforming the "Status" feature from a character to a factor feature
df2$Status <- as.factor(df2$Status)

# Checking how the "Status" feature is now
summary(df2$Status)
```
```{r}
# Status feature boxplot
df2 %>% ggplot(aes(x = Life.expectancy, fill = Status)) + geom_boxplot() + xlab("Life Expectancy")
# Status feature histogram
df2 %>% ggplot(aes(x = Life.expectancy, fill = Status)) + geom_histogram(bins = 15) + facet_wrap(~Status) + xlab("Life Expectancy")
```
### The summary shows there are two factors in the "Status" feature: Developed and Developing.  The histograms do not show any strong evidence against normality.  With that in mind, I will perform a Welch's t-test to see if there is a difference in the mean Life Expectancy between the Developed status vs. the Developing status.

```{r}
Developed <- df[df$Status == "Developed",]$Life.expectancy
Developing <- df[df$Status == "Developing",]$Life.expectancy
t.test(Developed, Developing)
```
### The t-test results indicate that there is overwhelming evidence that the mean life expectancy from developed countries is not the same as the mean life expectancy of developing country.  The results also shows, with 95% confidence, that the mean life expectance from developed countries is approximately between 11.47 to 12.47 years longer than the mean life expectance of the developing countries.  These results are great to keep in mind when picking the features to implement in the model.

### In order to use the Status feature, I will create a dummy feature called "Developed" to replace it with a numeric feature.  The values in the "Developed" feature will be 1 if the "Status" is "Developed" or 0 if the "Status" is Developing.
```{r}
# Creating a "Developed" feature
df2$Developed <- ifelse(df$Status == "Developed", 1, 0)
# Dropping the "Status" feature
df2 <- df2 %>% select(!Status)
```

### The knn regression model will take the average value of the k nearest neighbors.  To find features that will produce neighbors that can best predict the life expectancy, I will use the correlation value each feature has to the Life.expectancy.  The stronger the correlation, the better.

```{r}
# library with ggpairs
library(GGally)

# Creating a new dataframe just to be able to change the features' titles
df3 <- df2

# Changing the features' titles just so that they can best fit the ggpairs graph
names(df3) <- c("Life expectancy",  "Adult Mortality", "infant deaths", 
                "percentage expenditure", "Measles", "BMI", "under five deaths", 
                "Polio", "Diphtheria", "HIV AIDS", "thinness 10 - 19 years", 
                "thinness 5 - 9 years", "Developed")

# Creating the ggpairs graph
ggpairs(df3, labeller = label_wrap_gen(8))
```

```{r}
# Creating a correlation dataframe
c <- data.frame(cor(df2))
# Checking the correlation dataframe where the absolute correlation from Life.expectancy is above 0.4
c[abs(c$Life.expectancy) > 0.4,]
```

```{r}
# Setting a seed to make it replicable
set.seed(1)

# Making an 80-20 train-test split
# Randomly selecting the indexes for the training set 
train.index <- sample(dim(df2)[1], dim(df2)[1] * 0.80)

# Creating the training set using the train.index
train <- df2[train.index,]

# Creating the testing set
test <- df2[-train.index,]
```


```{r}
# library with trainControl, train, RMSE, and MAE
library(caret)

# Creating a control variable to implement a 5 fold cross validation
fitControl <- trainControl(method = "cv", number = 5)

# Creating a knn model that will choose between 1 to 15 values for the "k" neighbors from which to take the mean.
knn.model <- train(Life.expectancy ~ Developed + Adult.Mortality + BMI + HIV.AIDS + thinness..1.19.years + thinness.5.9.years, 
                    method = "knn", tuneGrid = expand.grid(k = 1:15), data = train, trControl = fitControl)

# Checking the model
knn.model

# Creating a test.x1 data set with only the variable used in the creation of the model, without the response variable.
test.x <- test %>% select(!Life.expectancy)
test.x <- test %>% select(Developed, Adult.Mortality, BMI, HIV.AIDS, thinness..1.19.years, thinness.5.9.years)

# Making predictions with the testing data
knn.pred <- predict(knn.model, test.x)

RMSE(knn.pred, test$Life.expectancy)
MAE(knn.pred, test$Life.expectancy)
```

```{r}
# library with tree
library(tree)

# Tree Model
tree.model <- tree(Life.expectancy ~ ., data = train)
summary(tree.model)
```

```{r}
# Plotting the tree model and its nodes
plot(tree.model)
text(tree.model)
```

```{r}
# Building a dataset with only the predictors from the test dataset
test.x <- test %>% select(-Life.expectancy)
# Making the prediction using the regression tree model
tree.pred <- predict(tree.model, test.x)

# Checking the RMSE and MAE from the predicted model
RMSE(tree.pred, test$Life.expectancy)
MAE(tree.pred, test$Life.expectancy)
```